---
title: "Project Checkpoint 3"
author: "Kenny, Cole, Alden"
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## The Data

The following link displays the Kaggle source for one of the csvs used in this lab: [NBA Data from 2020-2023](https://www.kaggle.com/datasets/ahmedalmadami/2020-2023-nba-data/data)

This dataset contains NBA player performance statistics from the 2020-2023 NBA seasons. Each row shows the stats for a particular player in a particular season between 2020-2023. There are a number of different variables that fall into following categories:

-   Player Context: nationality, college, name, height, weight, age
-   Injury Statistics: games missed, cash earned while out, type of injury
-   Team Context: wins, losses, team, division
-   A wide array of advanced game performance stats.

The following link displays the Kaggle source for the second csv used in this lab: [NBA Salary Data 2023](https://www.kaggle.com/code/rizal1015/nba-salaries?select=nba_salaries.csv)

This csv contains information primarily on each individual players NBA salary and player performance statistics during the 2023 season. This data differs from the previous dataset in that player position, salary, and different player performance statistics are being used.

These datasets are collectively useful to analyze player productivity, salary trends, compare performance between seasons, and identify statistical trends among NBA players. In addition, using these two sources will allow students to practice joining and cleaning datasets in order to solve problems throughout this lab.

## Read in Datasets + Load Packages

```{r}
# Load in Necessary Packages
library(tidyverse)
library(knitr)
```

```{r}
# Read in the Datasets
nba_data <- read_csv("nba_data.csv")
nba_salary <- read_csv("nba_salaries.csv")
```

## Question 1

**1. Join the 2023 NBA salary dataset with the "nba_data" dataset. Keep only the player name, position, and salary columns from the NBA salary data set to avoid advanced statistics redundancy. Keep only players who appear in both datasets and ensure the data only comes from the 2023 season.**

```{r}
# Code for Question 1
nba_combined <- nba_salary |>
  select(`Player Name`, Position, Salary) |>
  inner_join(nba_data,
             by = c("Player Name" = "Player")) |> 
  filter(Season == 2023)
```

## Question 2

Now that we have the combined dataset, we would like to analyze high performing positions in the NBA to see if their position is currently being underpaid or overpaid. Win shares is a valuable NBA statistic that estimates the number of wins that a player contributed to their teams success. Since this variable has the potential to be influenced by players who play more minutes throughout a season, we can use the `WS/48` variable which produces the win shares estimated per 48 minutes which reduces the influence of a player playing more minutes. 

**2. Using the combined dataset we just made, calculate the overall median NBA salary. Then, create a nicely formatted table with one row for each position containing the following variables:**

-   median Win Shares per 48 minutes (WS/48)
-   median salary by position
-   the difference between the position median salary and the overall NBA median salary
-   the number of players in each position

**Only include positions with more than 10 players, and order the rows from highest to lowest median Win Shares.**

:::callout-tip

:::

```{r}
nba_combined |>
  mutate(tot_median = median(Salary.x, na.rm = T)) |> 
  group_by(Position) |>
  summarize(`median_WS/48` = median(`WS/48`, na.rm = T),
            median_position = median(Salary.x, na.rm = T),
            salary_difference = median_position - mean(tot_median),
            players = n()) |>
  filter(players > 10) |> 
  arrange(desc(`median_WS/48`)) |>
  kable()
```

## Question 3

In Question 2, we compared positions using median salary and median WS/48 separately, but those summaries don't directly tell us how efficiently teams are spending money on individual players. A player who produces high win shares on a small contract represents much better value than a similarly productive player on a max contract. To explore this, we can build a reusable function that summarizes player-level value within a single position, making it easy to compare any two positions side-by-side.

**3. Write a function that takes in a position and returns a summary table for players in that position. The table should include each player’s name, salary, WS/48, VORP, games missed, and a new variable called salary_per_WS48, calculated as salary divided by WS/48. Use the function to compare two different positions from Question 2 and determine which position appears to provide better value relative to salary.**

# Code for Question 3
```{r}
position_value <- function(pos_name) {
  nba_combined |>
    filter(Position == pos_name, !is.na(`WS/48`), `WS/48` > 0) |>
    mutate(
      salary_per_WS48 = Salary.x / `WS/48`
    ) |>
    select(
      `Player Name`,
      Position,
      Salary.x,
      `WS/48`,
      VORP,
      `Games Missed`,
      salary_per_WS48
    ) |>
    arrange(salary_per_WS48) |>
    kable(
      col.names = c(
        "Player",
        "Position",
        "Salary",
        "WS/48",
        "VORP",
        "Games Missed",
        "Salary per WS/48"
      ),
      digits = 2
    )
}

position_value("PG")
position_value("C")
```


## Question 4

**4. While the tables from Question 3 are useful, a visualization can make value comparisons across positions much clearer. Using the combined dataset from Question 1, create a nicely formatted ggplot that compares WS/48 and Salary across positions. Your graph should: **

- Display all positions with more than 10 players (from Question 2)
- Make it visually clear which positions tend to produce more win shares per dollar spent
- Include informative axis labels, a title, and a caption noting the data source
- Apply a clean theme (e.g., theme_minimal() or theme_bw())


```{r}
#| label: q4-value-plot
#| fig-width: 9
#| fig-height: 6

nba_combined |>
  group_by(Position) |>
  filter(n() > 10) |>
  ungroup() |>
  filter(!is.na(`WS/48`), !is.na(Salary.x)) |>
  ggplot(aes(x = `WS/48`, y = Salary.x / 1e6, color = Position)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_y_continuous(labels = scales::label_dollar(suffix = "M")) +
  labs(
    title = "Salary vs. Win Shares per 48 Minutes by Position",
    subtitle = "Flatter trend lines indicate positions where higher salaries are not strongly tied to better production",
    x = "Win Shares per 48 Minutes (WS/48)",
    y = "Salary (2023 Season)",
    color = "Position",
    caption = "Source: Kaggle NBA datasets (2020–2023)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 10),
    legend.position = "bottom"
  )
```

Based on the graph, point guards (PG) appear to provide the best value relative to salary and tend to cluster in the higher WS/48 range without commanding the highest salaries, meaning teams are getting strong per-minute production at a lower cost. Centers (C), by contrast, show a wider salary spread without a proportional increase in WS/48, suggesting some centers are overpaid relative to their win-share contribution. This is consistent with the conclusion from Question 3, where PGs had a lower salary-per-WS/48 than Cs.

## Question 5

While much of NBA analysis focuses on performance and salary, player availability is just as important — a star player who frequently misses games provides less value to their team than their per-game stats might suggest. Our dataset includes an `Injury` column containing free-text descriptions of each player's injury that season (e.g., "sprained left ankle," "torn ACL"). By extracting keywords from these descriptions, we can group similar injuries together and analyze how different injury types impact player availability.

**5. Create a new variable called `injury_category` that classifies each player's injury as `"Knee"`, `"Ankle"`, `"Back"`, `"Hamstring"`, `"Other"`, or `"None"` based on keywords found in the `Injury` text. Then, create a nicely formatted table summarizing the number of players and the median number of games missed for each injury category, ordered from most to fewest median games missed. Which type of injury appears to be the most costly in terms of player availability?**

````{r}
#| label: q5-injury-categories

nba_combined |>
  mutate(
    injury_lower = str_to_lower(Injury),
    injury_category = case_when(
      is.na(Injury) | Injury == ""           ~ "None",
      str_detect(injury_lower, "knee|acl|mcl|meniscus") ~ "Knee",
      str_detect(injury_lower, "ankle")      ~ "Ankle",
      str_detect(injury_lower, "back|spine|lumbar")     ~ "Back",
      str_detect(injury_lower, "hamstring")  ~ "Hamstring",
      TRUE                                    ~ "Other"
    )
  ) |>
  group_by(injury_category) |>
  summarize(
    players = n(),
    median_games_missed = median(`Games Missed`, na.rm = TRUE)
  ) |>
  arrange(desc(median_games_missed)) |>
  kable(
    col.names = c("Injury Category", "Number of Players", "Median Games Missed"),
    caption = "Player Availability by Injury Type (2023 Season)"
  )
````

Based on the table, knee injuries appear to be the most costly in terms of player availability — players with knee injuries miss substantially more games on average than players with injuries to other body parts. This makes intuitive sense given that serious knee injuries (such as ACL or meniscus tears) often require surgery and lengthy rehabilitation, whereas ankle and hamstring injuries tend to heal more quickly. Teams may want to weigh injury history, especially knee-related injuries more heavily when evaluating player contracts.



## Question 6

So far, we've used the `position_value()` function from Question 3 to look at one position at a time. To find the most valuable contracts across the entire league, we'd need to apply the function to every qualifying position and combine the results into a single ranking. Rather than calling the function manually for each position, we can use iteration with `map_dfr()` from the `purrr` package (part of the tidyverse) to apply the function to all positions at once and stack the resulting tables together.

**6. Using iteration, apply a player-level value function to every position with more than 10 players (from Question 2). Combine the results into a single data frame, then display the 10 players league-wide with the lowest `salary_per_WS48` in a nicely formatted table. Which players represent the best contract value in the league?**

```{r}
#| label: q6-best-value-league

#Modified version
position_value_df <- function(pos_name) {
  nba_combined |>
    filter(Position == pos_name, !is.na(`WS/48`), `WS/48` > 0) |>
    mutate(salary_per_WS48 = Salary.x / `WS/48`) |>
    select(
      `Player Name`,
      Position,
      Salary.x,
      `WS/48`,
      VORP,
      `Games Missed`,
      salary_per_WS48
    )
}

#Find positions with more than 10 players
positions_to_check <- nba_combined |>
  count(Position) |>
  filter(n > 10) |>
  pull(Position)

#Iterate the function over every qualifying position, stacking results
all_positions_value <- map_dfr(positions_to_check, position_value_df)

#Display the top 10 best-value players in the dataset
all_positions_value |>
  arrange(salary_per_WS48) |>
  slice_head(n = 10) |>
  kable(
    col.names = c(
      "Player",
      "Position",
      "Salary",
      "WS/48",
      "VORP",
      "Games Missed",
      "Salary per WS/48"
    ),
    digits = 2,
    caption = "Top 10 Best Contract Values in the NBA (2023 Season)"
  )
```

The resulting table reveals the players providing the most win-share production per dollar of salary across the entire league. Many of these players tend to be on rookie-scale or minimum contracts, since their salaries are capped by collective bargaining rules even when their on-court production is strong. This pattern is consistent with the position-level findings from Questions 3 and 4 — younger, less expensive players (often point guards still on their rookie deals) frequently top the value rankings, while veterans on max contracts rarely appear, even when their production is excellent.

## Question 7

Summary statistics like the median WS/48 we calculated in Question 2 are useful, but they only give us a single number — they don't tell us how *stable* that estimate is. If we had sampled a slightly different set of point guards, would the median look noticeably different? Bootstrap resampling helps us answer this by repeatedly drawing samples (with replacement) from our data, recalculating the statistic each time, and examining the distribution of those resampled values. This gives us a sense of the uncertainty around our estimate and lets us build a confidence interval for the true median.

**7. Use bootstrap resampling to estimate the uncertainty around the median WS/48 for point guards. Draw 1,000 bootstrap resamples (with replacement) from the point guard WS/48 values, compute the median of each resample, and create a nicely formatted histogram of the bootstrap distribution. Add vertical lines showing the 95% confidence interval (the 2.5th and 97.5th percentiles) and the overall league-wide median WS/48 from Question 2. Does the league-wide median fall inside the point guard confidence interval, and what does that suggest about how point guards compare to the rest of the league?**

```{r}
#| label: q7-bootstrap-pg-median
#| fig-width: 9
#| fig-height: 6

# Set a seed for reproducibility
set.seed(123)

# Pull out point guard WS/48 values
pg_ws48 <- nba_combined |>
  filter(Position == "PG", !is.na(`WS/48`)) |>
  pull(`WS/48`)

# Bootstrap: 1,000 resamples, computing the median of each
boot_medians <- map_dbl(
  1:1000,
  ~ median(sample(pg_ws48, size = length(pg_ws48), replace = TRUE))
)

# 95% confidence interval from the bootstrap distribution
boot_ci <- quantile(boot_medians, probs = c(0.025, 0.975))

# League-wide median WS/48 (matches Question 2)
nba_median_ws48 <- median(nba_combined$`WS/48`, na.rm = TRUE)

# Plot the bootstrap distribution
tibble(boot_median = boot_medians) |>
  ggplot(aes(x = boot_median)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white",
    alpha = 0.85
  ) +
  geom_vline(
    xintercept = boot_ci,
    color = "darkorange",
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = nba_median_ws48,
    color = "firebrick",
    linetype = "solid",
    linewidth = 1
  ) +
  labs(
    title = "Bootstrap Distribution of Median WS/48 for Point Guards",
    subtitle = "Dashed orange lines show the 95% confidence interval; solid red line shows the league-wide median",
    x = "Bootstrapped Median WS/48",
    y = "Frequency",
    caption = "Source: Kaggle NBA datasets (2020–2023) | 1,000 bootstrap resamples"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )

# Print the numerical results
boot_ci
nba_median_ws48
```

The 95% confidence interval shows the range of plausible values for the true median point guard WS/48, given our sample data. If the overall league-wide median falls *outside* the point guard confidence interval, it suggests point guards have a meaningfully different typical WS/48 than the rest of the league, supporting the conclusion from Question 2 that point guards have a higher-than-average median WS/48. If the league-wide median falls *inside* the interval, point guards' median isn't statistically different from the league as a whole at this sample size, and the differences we observed in Question 2 may not be as conclusive as the table suggested. The bootstrap distribution gives us a visualization of the data which is more helpful than a single median value.






