---
title: "Lab: Basketball Data"
author: "Kenny, Cole, Alden"
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## The Data

The following link displays the Kaggle source for one of the csvs used in this lab: [NBA Data from 2020-2023](https://www.kaggle.com/datasets/ahmedalmadami/2020-2023-nba-data/data)

This dataset contains NBA player performance statistics from the 2020-2023 NBA seasons. Each row shows the stats for a particular player in a particular season between 2020-2023. There are a number of different variables that fall into following categories:

-   Player Context: nationality, college, name, height, weight, age
-   Injury Statistics: games missed, cash earned while out, type of injury
-   Team Context: wins, losses, team, division
-   A wide array of advanced game performance stats.

The following link displays the Kaggle source for the second csv used in this lab: [NBA Salary Data 2023](https://www.kaggle.com/code/rizal1015/nba-salaries?select=nba_salaries.csv)

This csv contains information primarily on each individual players NBA salary and player performance statistics during the 2023 season. This data differs from the previous dataset in that player position, salary, and different player performance statistics are being used.

These datasets are collectively useful to analyze player productivity, salary trends, compare performance between seasons, and identify statistical trends among NBA players. In addition, using these two sources will allow students to practice joining and cleaning datasets in order to solve problems throughout this lab.

## Read in Datasets + Load Packages

```{r}
# Load in Necessary Packages
library(tidyverse)
library(knitr)
library(broom)
library(kableExtra)
library(scales)

# Read in the Datasets
nba_data <- read_csv("nba_data.csv")
nba_salary <- read_csv("nba_salaries.csv")
```

## Question 1

**1. Join the 2023 NBA salary dataset with the "nba_data" dataset. Keep only the player name, position, and salary columns from the NBA salary data set to avoid advanced statistics redundancy. Keep only players who appear in both datasets and ensure the data only comes from the 2023 season.**

```{r}
#| label: Q1-Data-Join
# Code for Question 1

```

## Question 2

Now that we have the combined dataset, we would like to analyze high performing positions in the NBA to see if their position is currently being underpaid or overpaid. Win shares is a valuable NBA statistic that estimates the number of wins that a player contributed to their teams success. Since this variable has the potential to be influenced by players who play more minutes throughout a season, we can use the `WS/48` variable which produces the win shares estimated per 48 minutes which reduces the influence of a player playing more minutes. 

**2. Using the combined dataset we just made, calculate the overall median NBA salary. Then, create a nicely formatted table with one row for each position containing the following variables:**

-   median Win Shares per 48 minutes (WS/48)
-   median salary by position
-   the difference between the position median salary and the overall NBA median salary
-   the number of players in each position

**Only include positions with more than 10 players, and order the rows from highest to lowest median Win Shares.**

::: {.callout-tip}
Use KableExtra or gt styling in order to improve the appearance of your tables. In addition, use the package 'scales' to format dollar output for relevant columns. 
:::

```{r}
#| label: Q2-Median-Table

# Code for Question 2
```

## Question 3

In Question 2, we compared positions using median salary and median WS/48 separately, but those summaries don't directly tell us how efficiently teams are spending money on individual players. A player who produces high win shares on a small contract represents much better value than a similarly productive player on a max contract. To explore this, we can build a reusable function that summarizes player-level value within a single position, making it easy to compare any two positions side-by-side.

**3. Write a function that takes in a position and returns a summary table for players in that position. The table should include each player’s name, salary, WS/48, VORP, games missed, and a new variable called salary_per_WS48, calculated as salary divided by WS/48. Use the function to compare two different positions from Question 2 and determine which position appears to provide better value relative to salary.**

::: {.callout-tip}
To avoid outputting the entire list, insert another argument function to only output the first 5 NBA players.
:::

```{r}
#| label: Q3-Position-Function

# Code for Question 3
```

## Question 4

**4. While the tables from Question 3 are useful, a visualization can make value comparisons across positions much clearer. Using the combined dataset from Question 1, create a nicely formatted scatterplot that compares WS/48 and Salary across positions. Your graph should: **

- Display all positions with more than 10 players (from Question 2)
- Make it visually clear which positions tend to produce more win shares per dollar spent
- Include informative axis labels, a title, and a caption noting the data source
- Apply a clean theme (e.g., theme_minimal() or theme_bw())


```{r}
#| label: Q4-Salary-Visualization

# Code for Question 4
```

*Answer question here*

## Question 5

While much of NBA analysis focuses on performance and salary, player availability is just as important. Our dataset includes an `Injury` column containing descriptions of each player's injury that season (e.g., "sprained left ankle," "torn ACL"). By extracting keywords from these descriptions, we can group similar injuries together and analyze how different injury types impact player availability.

**5. Create a new variable called `injury_category` that classifies each player's injury as `"Knee"`, `"Ankle"`, `"Back"`, `"Hamstring"`, `"Other"`, or `"None"` based on keywords found in the `Injury` text. Then, create a nicely formatted table summarizing the number of players and the median number of games missed for each injury category, ordered from most to fewest median games missed. Which type of injury appears to be the most costly in terms of player availability?**

::: {.callout-tip}
Use `str_to_lower()` before searching injury text so that words like "Knee", "knee", and "KNEE" are treated the same.
:::

````{r}
#| label: Q5-Injury-Categories

# Code for Question 5
````

*Answer question here*

## Question 6

So far, we've used the `position_value()` function from Question 3 to look at one position at a time. To find the most valuable contracts across the entire league, we'd need to apply the function to every qualifying position and combine the results into a single ranking. Rather than calling the function manually for each position, we can use iteration with `map_dfr()` from the `purrr` package (part of the tidyverse) to apply the function to all positions at once and stack the resulting tables together.

**6. Using iteration, apply a player-level value function to every position with more than 10 players (from Question 2). Combine the results into a single data frame, then display the 10 players league-wide with the lowest `salary_per_WS48` in a nicely formatted table. Which players represent the best contract value in the league?**

::: {.callout-tip}
Create a function that returns a data frame, not a `kable()` table. This makes it easier to combine results across positions using `map_dfr()`.
:::

```{r}
#| label: Q6-Best-Value-Function

# Code for Question 6
```


```{r}
#| label: Q6-Best-Value-Function

# Code for Question 6 Continued
```

*Answer question here*

## Question 7

**7a. Salary is often influenced by player performance, but it is not always clear which statistics matter most. Using the combined 2023 dataset, fit a linear regression model predicting salary using WS/48, VORP, and Games Missed as explanatory variables. Display the model summary comment on which variable appears to have the strongest relationship with salary?**

```{r}
#| label: Q7a-Model

# Code for Question 7a
```

*Answer question here*

**7b. Conduct model diagnostics to check that the conditions for linear regression that we discussed in class are met (L.I.N.E.). Comment on whether the `salary_model` meets the appropriate L.I.N.E conditions. **

::: {.callout-tip}
You may need multiple plots or output to check all diagnostics. You should be using techniques that we learned in class - not base R functions.
:::

```{r}
#| label: Q7b-Model-Diagnostics

# Code for Question 7b
```

*Answer question here*

## Question 8

**Question 8: Using the combined 2023 dataset, classify each player as "Low Salary", "Middle Salary", or "High Salary" based on salary percentiles. Then, summarize the median WS/48, median VORP, median salary, and number of players in each salary category. Which salary group appears to provide the best overall performance value?**

```{r}
#| label: Q8-Salary-Categories

# Code for Question 8
```

*Answer question here*

## Question 9

**Question 9: Simulate 1,000 random samples of 25 players from the 2023 combined dataset. For each sample, calculate the mean salary. Then, create a histogram of the simulated mean salaries. How does the average simulated sample salary compare to the true mean salary of all players in the dataset?**

```{r}
#| label: Q9-Salary-Simulation

# Code for Question 9
```

```{r}

# Code for Question 9 Continued
```

*Answer question here*

## Question 10

**Question 10: Using the combined 2023 dataset, create a new variable called performance_tier that classifies players as "Low", "Average", or "High" performers based on their VORP values. Then, create a faceted ggplot showing the relationship between Salary and WS/48 for each performance tier. What does the visualization suggest about whether higher-paid players are always more efficient?**

::: {.callout-tip}
Use the qnorm() function in R to classify players performance based on VORP values. Let a VORP value in the 33rd percentile be classified as "Low", between the 33rd and 67th percentile be classified as "Average" and anything higher be classified as "High". 
:::

```{r}
#| label: Q10-Performance-Tier

# Code for Question 10
```

*Answer question here*
